---
title: "Exploratory Analysis Project"
author: "Alicia Quinteros"
date: "2020-12-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

##### The two datasets that I choose for this project are: Energy Census and Economic Data in the United States from 2010-2014 and Pollution levels in the US from 2000-2016. The Energy and Census data set contains the region, total energy consumption per year, total energy production, and total expidentures per year in each state. Moreover, the Pollution dataset contains four different pollutants: Methane, Ozone, Nitrogen Dioxide, and Sulfur Dioxide found in the different states of the US. I obtained these datasets from Kaggle where the pollution data was aquired through the US EPA and the Energy dataset used the Census website and the EIA for the energy information. 

##### By comparing the amount of energy produced and used in the US by the people, I would like to find out the correlation with the levels of pollution in the atmosphere in comparison to the usage of energy in each state. Looking into the different states and the different levels of energy consumption, we would be able to determine if population size plays a big role in how much pollution is generated due to the usage and creation of energy. I am interested in this topic because climate change will be a big factor in the upcoming years and as a way to further understand how much pollution is being generated.  

```{r cars}
library(tidyverse)
library(dplyr)
getwd()
library("readr")
Pollution <- read.csv("/stor/home/aeq242/SDS348_Fall2020/Data/pollution_us_2000_2016.csv")
Energy <- read.csv("/stor/home/aeq242/Energy Census and Economic Data US 2010-2014.csv")
```

## Tidying and Rearranging Datasets

```{r pressure, echo=FALSE}
untidy_pollution <- Pollution %>% pivot_wider(names_from = "NO2.1st.Max.Hour", values_from = "State")
untidy_pollution %>% pivot_longer(cols = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "12", "13", "14", "15", "16", "17", "18", "19", "20", "21", "22", "23"),names_to = "NO2.1st.Max.Hour", values_to = "State")

untidy_energy <- Energy %>% pivot_wider(names_from = "Great.Lakes", values_from = "State")
untidy_energy %>% pivot_longer(cols = c("0","1"), names_to = "Great.Lakes", values_to = "State")
```
My datasets, Energy and Pollution, were already tidy from the start by containing a variable that matched its respective observation in its own column. By using pivot_wider, I was able to untidy them by creating a wider dataset, which made the observations from one variable receive their own columns as if they were variables. Then the observations from the second variable set became data for the rows of these new columns. The unnecessary extra columns were then made tidy by using pivot_longer so that the columns for the observations received their own column.

## Joining and Merging
```{R}
EnergyandPollution <- Pollution %>% full_join(Energy, by = "State")
```

To combine my Pollution and Energy dataset, I used full_join and joined them by their common column which was State. There were 52 observations and 192 variables for the Energy dataset. For the Pollution data, there was a total of 1,746,661 observations and 29 variables. When combining both sets of data we get: 1,746,668 observations and 220 variables. No observations were dropped, all the data was retained! I used the code full_join to merge the datasets because my variables were not in the same position. Therefore, there were Na's that were introduced when there wasn't a match between the two datasets resulting in a slightly bigger observation size. 

## Wrangling
```{r, warning=FALSE}
FilteredData <- EnergyandPollution %>% filter(CO.Mean > 0 & str_detect(Date.Local, "2014") & TotalP2014 >= 1000000 & State!= "United State") %>% arrange(desc(TotalP2014), desc(CO.Mean)) %>% mutate(Combined.Gasses = NO2.Mean + O3.Mean +SO2.Mean +CO.Mean) %>% select(State, City, Date.Local, CO.Mean, TotalP2014, Combined.Gasses)
head(FilteredData)


EPMean <- EnergyandPollution %>% group_by(State) %>% summarize(mean_CO = mean(CO.Mean, na.rm = T), mean_NO2 = mean(NO2.Mean, na.rm=T), mean_O3 = mean(O3.Mean, na.rm = T), mean_SO2 = mean(SO2.Mean, na.rm = T), mean_TotalP2014 = mean(TotalP2014, na.rm = T), mean_TotalC2014 = mean(TotalC2014, na.rm = T), mean_CoalC2014 = mean(CoalC2014, na.rm = T), mean_CoalP2014 = mean(CoalP2014, na.rm = T), mean_NatGasC2014 = mean(NatGasC2014, na.rm = T), mean_FossFuelC2014 = mean(FossFuelC2014, na.rm = T))
EPMean

EnergyandPollution %>% group_by(State) %>% summarize(max_CO = max(CO.Mean, na.rm = T), max_NO2 = max(NO2.Mean, na.rm=T), max_O3 = max(O3.Mean, na.rm = T), max_SO2 = max(SO2.Mean, na.rm = T), max_TotalP2014 = max(TotalP2014, na.rm = T), max_TotalC2014 = max(TotalC2014, na.rm = T), max_CoalC2014 = max(CoalC2014, na.rm = T), max_CoalP2014 = max(CoalP2014, na.rm = T), max_NatGasC2014 = max(NatGasC2014, na.rm = T), max_FossFuelC2014 = max(FossFuelC2014, na.rm = T))


EnergyandPollution %>% group_by(State) %>% summarize(min_CO = min(CO.Mean, na.rm = T), min_NO2 = min(NO2.Mean, na.rm=T), min_O3 = min(O3.Mean, na.rm = T), min_SO2 = min(SO2.Mean, na.rm = T), min_TotalP2014 = min(TotalP2014, na.rm = T), min_TotalC2014 = min(TotalC2014, na.rm = T), min_CoalC2014 = min(CoalC2014, na.rm = T), min_CoalP2014 = min(CoalP2014, na.rm = T), min_NatGasC2014 = min(NatGasC2014, na.rm = T), min_FossFuelC2014 = min(FossFuelC2014, na.rm = T))

EnergyandPollution %>% group_by(State) %>% summarize(median_CO = median(CO.Mean, na.rm = T), median_NO2 = median(NO2.Mean, na.rm=T), median_O3 = median(O3.Mean, na.rm = T), median_SO2 = median(SO2.Mean, na.rm = T), median_TotalP2014 = median(TotalP2014, na.rm = T), median_TotalC2014 = median(TotalC2014, na.rm = T), median_CoalC2014 = median(CoalC2014, na.rm = T), median_CoalP2014 = median(CoalP2014, na.rm = T), median_NatGasC2014 = median(NatGasC2014, na.rm = T), median_FossFuelC2014 = median(FossFuelC2014, na.rm = T))

EnergyandPollution %>% group_by(State) %>% summarize(distinct_CO = n_distinct(CO.Mean, na.rm = T), distinct_NO2 = n_distinct(NO2.Mean, na.rm=T), distinct_O3 = n_distinct(O3.Mean, na.rm = T), distinct_SO2 = n_distinct(SO2.Mean, na.rm = T), distinct_TotalP2014 = n_distinct(TotalP2014, na.rm = T), distinct_TotalC2014 = n_distinct(TotalC2014, na.rm = T), max_CoalC2014 = n_distinct(CoalC2014, na.rm = T), distinct_CoalP2014 = n_distinct(CoalP2014, na.rm = T), distinct_NatGasC2014 = n_distinct(NatGasC2014, na.rm = T), distinct_FossFuelC2014 = n_distinct(FossFuelC2014, na.rm = T))

EnergyandPollution %>% group_by(State, City) %>% summarize(mean_CO = mean(CO.Mean, na.rm = T), max_CO = max(CO.Mean, na.rm= T))

library(tibble)
untidycor <- EnergyandPollution %>% filter(CO.Mean > 0 & str_detect(Date.Local, "2014") & TotalP2014 >= 1000000 & State!= "United State") %>% arrange(desc(TotalP2014), desc(CO.Mean))%>% select(State, City, Date.Local, CO.Mean, TotalP2014, CoalP2014, CoalC2014, FossFuelC2014, NO2.Mean, O3.Mean, SO2.Mean, NatGasC2014)
CorEP <- untidycor %>% select_if(is.numeric) %>% cor(use = "pair") 
cordata <- CorEP %>% as.data.frame %>% rownames_to_column("var1") %>% pivot_longer(-1, names_to = "var2", values_to = "correlation")
cordata
```

In the combined dataset, EneryandPollution, we want to look at a similar time period for the individual states. Since we are using the Total Energy Produced in 2014, we filtered out the year 2014 from Date.Local from pollution as a way to compare the levels of monoxide in relation to the amount of energy created. However, we remove the observation "United States" so we are only left with individual observations by state. Additionally, we used the code "arrange" in descending order to see which state is producing the most Energy and which state has the most Carbon Monoxide levels, resulting in Texas! But, we want to compare the correlation of the greenhouse gasses, as well, to the total production of energy. We do this by using the code mutate and add the means of all the gasses to create the column, Combined.Gasses. Now, we are able to select the data we will be focusing on by using select. 

Each state had multiple observations of means for each gas (NO2, O3, SO2, and CO) and I wanted to take the average of all the means for each state to end up with a single value that represented the mean of the state altogether. However, I also wanted to compare how this correlated with energy consumption, energy production, coal production, coal consumption, and natural gas consumption. To accomplish this, I used group_by to link the states with their respective observation and summarize to take the : mean, max, min, median and distinct of each of these variables. Looking through our data, we can see that NO2 had the highest mean, median, and maximum for all states. The gas with minimum levels for all states was SO2. Each state had many distinct observations for the gasses. Lastly, for our correlation test, we had to filter certain variables in order for the code to run. This is because, our original dataset was too large, with multiple numeric variables that were unnecessary to run in a correlation such as price, country code, deaths and rebirths. As a result, we found that there is a strong correlation with Fossil Fuel consumption, Energy Production and Natural Gas consumption. But the levels of gasses in the atmosphere do not directly correlate. 


## Visualization
```{r, warning=FALSE}
library(ggplot2)
cordata %>% ggplot(aes(var1,var2,fill= correlation)) + geom_tile() + geom_text(aes(label = round(correlation,2))) + xlab("") + ylab("") + coord_fixed() + scale_fill_gradient2(low = "red", mid = "white", high = "blue") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) 

smalldata <- EPMean %>% filter(mean_CO > 0 & mean_TotalP2014 >= 1000000 & State!= "United State" & State!="District of Columbia" & State!="Mississippi" & State!="Montana" & State!="Nebraska" & State!="Vermont" & State!="West Virginia") %>% arrange(desc(mean_TotalP2014), desc(mean_CO)) %>% select(State, mean_CO, mean_TotalP2014, mean_TotalC2014, mean_NatGasC2014, mean_NO2, mean_O3, mean_SO2) 

ggplot(smalldata, aes(mean_TotalP2014, mean_CO)) + geom_point(aes(color = State)) + geom_smooth(method = "lm") + ggtitle("Energy Production vs Carbon Monoxide Mean") + xlab("Total Energy Production 2014 (billion BTU)") + ylab("Carbon Monoxide Mean (parts per million)")

ggplot(smalldata, aes(x= mean_SO2, mean_CO, fill = State)) + geom_bar(stat = "summary", position = 'dodge') + scale_x_continuous(breaks = seq(0,7,1))+ggtitle("Sulfur Dioxide Mean vs Carbon Monoxide Mean") + xlab("Sulfur Dioxide Mean (parts per billion)") + ylab("Carbon Monoxide Mean (parts per million)")
```

Creating a correlation Heatmap, we can see the negative and positive correlation in our variables. As we can see, the production of Energy in 2014 does not seem to correlate with the levels of gasses in the atmosphere but positively correlate with the consumption of Fossil Fuels, Coal and Natural gasses by having values of 0.5 or greater. We know this by comparing their values from the scale 1.0 to -1.0, where 1 means it correlates completely and -1 means it correlates in opposite directions. Thus, using our heatmap we can see how the level of gasses correlate to one another, the mean of Carbon Monoxide positively correlating at 0.66 with the mean of Nitrogen Dioxide. Looking at the Total Production of Energy in 2014 in comparison to the Mean Carbon Monoxide, there is an opposite relationship of -0.27. This unexpected correlation might be due to the population size in that state, land size, where the sample was taken from, and other factors that influence the atmosphere. 

In our scatterplot, we compared the Total Energy Production in 2014 to the Mean Carbon Monoxide levels in each state. Because our original correlation dataset was still too large to form a coherent plot, I used the mean of each state to only get a single observation per state rather than having multiple states with many numeric observations. Thus, our plot shows how there is a negative trend line, indicating that the level of carbon monoxide in the atmosphere is negatively related to the production of energy. However, we can see that some states do have an extremely high production of energy and levels of monoxide (Texas) but they are considered outliers.  

Our third plot we compared the level of Carbon Monoxide to the level of Sulfur Dioxide as a way to compare the relationship between two gasses. The bar graph indicates that there is no discernible relationship between the gasses per state in the atmosphere. This means that because there is a high level of Carbon Monoxide in the atmopshere for that state, does not mean that there will be a high or low level of sulfur dioxide. For instance, California has a low level of Sulfur Dioxoide but one of the highest level of Carbon Monoxide. Alaska differs from California where it has a greater amount of Sulfur Dioxide and Carbon Monoxide. 

## Dimensionality Reduction  
```{r, warning=FALSE}

library(GGally)
library(cluster)
library(plotly)
clustereddata <- smalldata %>% select(-State)
pam1 <- clustereddata %>% scale %>% pam(5)
pam1
clustpam <- clustereddata %>% mutate(cluster = as.factor(pam1$clustering))

sil_width<-vector()
for(i in 2:10){
  pam_fit <- pam(clustereddata, diss = TRUE, k = i)
  sil_width[i] <- pam_fit$silinfo$avg.width 
  }
 ggplot()+geom_line(aes(x=1:10,y=sil_width))+scale_x_continuous(name="k",breaks=1:10)


plot(pam1, which = 2)  
 

clustpam%>%plot_ly(x= ~mean_CO, y = ~mean_TotalP2014, z = ~mean_NO2, color= ~cluster,
type = "scatter3d", mode = "markers") %>%
layout(autosize = F, width = 900, height = 400)

pam2 <- clustereddata %>% scale %>% pam(2)
clustpam1 <- clustereddata %>% mutate(cluster = as.factor(pam2$clustering))
ggpairs(clustpam1, columns = 1:3, aes(color = cluster))

```

In this section, I created a separated dataset by taking EPMean, which we created a while ago as a way to get a single observation per state rather than having multiple states with many numeric observations and selected only the numerical variables I will be focusing on by selecting States out and naming it clustereddata.Then I ran a cluster analysis using "pam" resulting in a table of Medoids, a Clustering vector, Objective function, and Available components. Next I checked the number of clusters in my data, using the silhouette method which resulted in the optimal number of clusters being 5. We created an empty vector and used a "for" function to create a pamfit for the cluster data. Then we were able to use ggplot with ten breaks in the x-axis to identify how many number of clusters was optimal. Furthermore, by running the code "pam1" and scaling it, I got an average silhouette width of 0.19 where there is no substantial structure found.

Lastly, by taking the 3 numerical variables in my dataset, mean_NO2, mean_TotalP2014, and mean_CO, I created a 3D visual with x,y, and z planes using plotly. This 3D model illustrates how there are only 5 clusters, indicating how there is no discernible pattern or cluster formation. It is difficult to interpret and could be due to the large geography of the US and where the atmospheric samples were taken. For instance, comparing the regions of the United States, there are many states with different elevations and population densities that contribute to the amount of gasses in the air. This would account as to why our goodness of fit test gave an average silhouette width of 0.19 since there was no substantial structure with these clusters. 


